---
title: "Introduction to R"
author: "Jessica Kosie"
date: "9/21/2018"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Goals 

By the end of this tutorial, you will:

+ have a better understanding of tidyverse `verbs`
+ understand the basics of graphing in ggplot2
+ understand the basics of writing in RMarkdown (content in a separate document)

The best way to do this tutorial is to walk through it slowly, executing each line and trying to understand what it does. You can execute a single line by placing your cursor somewhere in that line and hitting CMD+enter on mac or CTRL+enter on PC. 

First, set your working directory!

# 1 `tidyverse` review
  
If you haven't installed the tidyverse package, you'll need to run this command once:

`install.packages("tidyverse")`

Next, as always, we need to load the package with `library()`:

```{r}
# Load the packages (tell R that you want to use its functions)
library(tidyverse)
library(rio)
```

We're going to reread the data now, using `import`, which is from the `rio` library; it's not from `tidyverse()` per se, but its designed similarly and is easy to use.

```{r}
ps_data <- import("pragmatic_scales_data.csv")
```

## 1.1. Pipes

Pipes are a way to write strings of functions more easily. They bring the first argument of the function to the beginning. So you can write:

```{r}
sum(select(ps_data, age))

# becomes...

ps_data %>% 
  select(age) %>% 
  sum()
```

That's not very useful yet, but it really pays off with more and more steps. For example, let's say we want the average age of kids per condition, but first need to remove the repeated entries of the same kids:

```{r}
# This mess:
summarize(group_by(distinct(ps_data, sub_id, .keep_all = TRUE), condition), mean_age = mean(age, na.rm = TRUE))

# Becomes this:
ps_data %>% # take the ps_data, then...
  distinct(subid, .keep_all = TRUE) %>% # keep distinct subjects, then... 
  group_by(condition) %>% # group by condition, then...
  summarize(mean_age = mean(age, na.rm = TRUE)) # return the mean age (per group)

```

## 1.2. Filtering 

The verb `filter` can be used to extract *rows* from your dataset. You might do this to remove outliers, select a subpopulation, etc. `filter` is a verb (function) that takes a data frame as its first argument, followed by the **condition(s)** you want to filter on. 

So if you wanted to look only at kids between 2 and 3 year olds.

```{r}
ps_data %>%
  filter(age >= 2 & age =< 3)
```

## 1.3. Selecting
There are also times when you want to add or remove *columns*. You might want to remove columns to simplify the dataset. If you wanted to do that, the verb is `select`. 

```{r}
#get ONLY the subid, age, and correct columns 
ps_data %>%
  select(subid, age, correct) 
```

## 1.4. Mutating
Perhaps more useful is *adding columns*. You might do this perhaps to compute some kind of derived variable. `mutate` is the verb for these situations - it allows you to add a column. Let's mean-center age in the `ps_data` dataset. 
```{r}
ps_data <- ps_data %>%
  mutate(age_c = age - mean(age, na.rm = TRUE))

head(ps_data)
```
## 1.5. Arranging
Sometimes you might want to **sort** your data by the values in a particular column. For this, you'll use the verb `arrange`. Let's arrange our data by the `correct` column.

```{r}
ps_data %>% 
  arrange(age) 

ps_data %>% 
  arrange(desc(age))
```
## 1.6. Grouping and Summarising

Sometimes we want to describe a dataset at different levels. For instance, we can describe condition-level data in `ps_data` by combining two verbs: `group_by` and `summarize` . 
```{r}
# summarize without grouping
ps_data %>%
  summarise(correct_mean = mean(correct),
            correct_sd = sd(correct),
            n_observations = length(subid))

#grouped by age_group and condition
ps_data %>%
  group_by(condition) %>%
  summarise(correct_mean = mean(correct),
            correct_sd = sd(correct),
            n_observations = length(subid))
```

## Exercise 1a. 
>One of the most important analytic workflows for psychological data is to take some function (e.g., the mean) *for each participant* and then look at grand means and variability *across participant means*. This analytic workflow requires grouping, summarizing, and then grouping again and summarising again! First, create a table called `ps_sub_means` to get the mean `correct` for each `subid`. Hint: in addition to `subid`, group by `condition` in this summary to transfer those values to `ps_sub_means` (we'll use them in our next step). 

```{r}
ps_sub_means <- ps_data %>%
  group_by(condition, subid) %>%
  summarise(correct_mean = mean(correct), 
            n_observations = length(subid))

ps_sub_means
```

## Exercise 1b
>Next, create a table called `ps_means` where you use `ps_sub_means` as your data file and calculate the mean `correct` across `condition`.  

```{r}
ps_means <- ps_sub_means %>%
  group_by(condition) %>%
  summarise(correct_sd = sd(correct_mean),
            correct_mean = mean(correct_mean),
            n_subs = length(subid))
ps_means
```

# 2. Data tidying with `tidyr`
Credit: http://r4ds.had.co.nz/tidy-data.html#tidy-data-1

The two key verbs in **data tidying** are `spread` and `gather`. Each of these verbs relies on a key-value pair that contains a *key* that indicates *what* the information describes and a *value* that contains the actual information.

"Password: 0123456789"" is a key value pair. 0123456789 is the *value*, and it is associated with the *key* Password.

The `ps_data` we've been using is already *tidy*. Now, we'll look at some that are not! We'll use some datasets that are built into the tidyverse package. Each dataset has the same values of four variables, `country`, `year`, `population`, and `cases`, but each dataset organizes them in different ways.

Recall the "rules" for tidy data:
+ Each variable must have its own column.
+ Each observation must have it's own row.
+ Each value must have its own cell.

```{r}
?table1 #let's learn about the datset
table1 #this dataset is tidy!
```
## 2.1. Gathering & Joining Data

A common problem is a dataset where some column names are not names of variables, but *values* of a variable. Check out `table4a`. The column names `1999` and `2000` are not variables in our data, instead they represent values of the `year` variable, and each row represents two observations, not one. We need to *gather* these columns into a new pair of variables.

*gather* makes wide tables narrower and longer
```{r}
table4a
```
There are three questions we must answer:
1. Which columns do we want to gather? A: `1999` and `2000`
2. What is the name ofthe variable whose values form the column names (the *key*)? A: year
3. What is the name of the variable that is spread out over the cells (the *value*)? A: cases

We'll use these answers in the *gather* function:
```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")

```
> **Exercise 6a.** `table4b` contains information about the `population` variable. Let's *gather* that table as well. Type `table4b` to check it out before gathering. Your resulting table should have columns for `country`, `year`, and `population`.

```{r}
table4b #check it out


```
Note that table4a contained our `cases` data while table4b contained our `population` data. Everything else is the same. Let's join those two tables together using `left_join`. (see http://r4ds.had.co.nz/relational-data.html for more about joining data frames).
```{r}
tidy4a <- table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")

tidy4b <- table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")

left_join(tidy4a, tidy4b) #note that R tells you which columns were matched

```
## 7. Spreading Data
In contrast to *gather*ing, sometimes a single observation is scattered across multiple rows. Then, you'd want to use *spread* (which is the opposite of *gather*). In table2, a single observation is a country in a year (`type` is not a variable they are interested in including in their analyses), but each observation is spread across two rows.

*spread* makes long tables shorter and wider
```{r}
table2

```
Now, there are two questions we must answer:
1. Which column contains the variable names (i.e., the *key* column)? A: `type`
2. Which column contains the *values* (or the data from multiple variables)? A: `count`

We'll use these answers in the *spread* function:
```{r}
table2 %>% 
  spread(key = type, value = count)
```
> **Exercise 7a.** Let's play around with our ps_data. Make each `item` a unique variable. Use *spread* to reformat the data so that there is a unique column for each item. The values in each of the four `item` columns should indicate whether or not the subject got that particular item right or wrong (i.e., `correct` in ps_data). Hint: what is the *key*? What is the *value*? Do not save this as a new object.

```{r}


```
## 8. Graphing in ggplot2
For this section we're going to use another dataset that is built into R. It is called `iris`. Let's start by making a scatter plot of the relationship between Sepal.Length and Petal.Length.

Note: this is just the beginning! There are entire books on graphing in ggplot2! https://www.amazon.com/ggplot2-Elegant-Graphics-Data-Analysis/dp/331924275X/ref=as_li_ss_tl?ie=UTF8&linkCode=sl1&tag=ggplot2-20&linkId=4b4de5146fdafd09b8035e8aa656f300

```{r}
?iris #first, let's learn about the dataset

ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Petal.Length))

#this is the ggplot version of:
plot(iris$Sepal.Length, iris$Petal.Length)
```
Creation of a plot in ggplot2 begins with the function ggplot() which creates a coordinate system you can add data to. For example:
```{r}
ggplot(data = iris)
```
You then can add one or more layers to ggplot() to complete your graph. As above, we'll add a layer of points to our plot (creating a scatterplot). geom_point() is a *geom function*, each of which adds a different type of layer to a plot. Each *geom function* take a *mapping* argument that defines visual properties of the graph. 

Template for graphing in ggplot2:

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
  
```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Petal.Length))
```
Let's find out if the relationship between `Sepal.Length` & `Petal.Length` relates to the `Species` of iris. To visualize this, we are going to map `Species` to an *aesthetic* (or a visual property of one of the objects in our plot). Aesthetics include things like size, shape, or color of our points. Let's map the color of our points to the `Species` variable.
```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Petal.Length, color = Species))

#ggplot assigned each level of *color* to each unique value of `Species`. This is called *scaling*.
```
> **Exercise 8a.** Options for *aesthetics* include color, shape, size, and alpha. Create a scatter plot to visualize the relationship between `Sepal.Width` and `Petal.Width`. Add an aesthetic to visualize the effect of `Species`. Choose any aesthetic you'd like or play around with a few. What do they do? How might you use more than one aesthetic?

```{r}


```

> ProTip: We could also make separate graphs for each `Species` using `facet_wrap`. By passing a formula (data structure) to `facet_wrap`.

Here's how:
```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Petal.Length)) +
  facet_wrap(~ Species)
```
Maybe we'd prefer a line graph instead of a scatterplot to describe the `iris` data. In this case, we'd use a different *geom* (e.g., point, line, smooth, boxplot, bar). 

Let's make the same plot, using the smooth *geom* which fits a smoothed line to the data.
```{r}
ggplot(data = iris) +
  geom_smooth(mapping = aes(x = Sepal.Length, y = Petal.Length))

#loess is the default function for geom_smooth() 
# from http://www.statisticshowto.com/lowess-smoothing/: LOWESS (Locally Weighted Scatterplot Smoothing), sometimes called LOESS (locally weighted smoothing), is a popular tool used in regression analysis that creates a smooth line through a timeplot or scatter plot to help you to see relationship between variables and foresee trends.
```
We can add aesthetics to help visualize the data:
```{r}
ggplot(data = iris) +
  geom_smooth(mapping = aes(x = Sepal.Length, y = Petal.Length, linetype = Species))

ggplot(data = iris) +
  geom_smooth(mapping = aes(x = Sepal.Length, y = Petal.Length, linetype = Species, color = Species))
```
We can also layer multiple geoms in the same plot!
```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Petal.Length)) +
  geom_smooth(mapping = aes(x = Sepal.Length, y = Petal.Length))

#we could also have used global mappings
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point() +
  geom_smooth()

#finally, we can add an *aesthetic* to only one part of the graph
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point(mapping = aes(color = Species)) +
  geom_smooth()
```
> **Exercise 8b.** Plot the relationship between `Sepal.Width` and `Petal.Width`. As above, create lines overlaid on a scatter plot. For the points, use different colors for each `Species`. For the lines, use both different colors and line types for `Species`.

```{r}


```
Now let's try a bar plot. The bar chart below displays the total number of irises in the `iris` dataset, grouped by `Species`.
```{r}
ggplot(iris, aes(x = Species)) +
  geom_bar()
# not very interesting - it looks like there's an approximately equal number of irises for each species; note that geom_bar defaults to *count*. This would be more informative if we used another dataset that is built in to ggplot2 called `diamonds`.

ggplot(diamonds, aes(x = cut)) +
  geom_bar()
#this data set contains information about diamonds, and there are different numbers of diamonds at each level of `cut`
```
Let's get a more interesting plot for our `iris` dataset. We can look at the avearge `Sepal.Length` for each `Species` of iris.

First, we'll need to use the skills we've previously acquired to create a table of means that we'll then plot. We'll then use the argument stat = "identity" to map the mean value onto the y aesthetic.
```{r}
plot.data <- iris %>% 
  group_by(Species) %>% 
  summarise(mean = mean(Sepal.Length, na.rm=TRUE))

ggplot(plot.data, aes(x=Species, y=mean)) +
  geom_bar(stat="identity")

#more options:
ggplot(plot.data, aes(x=Species, y=mean, color=Species)) +
  geom_bar(stat="identity")

ggplot(plot.data, aes(x=Species, y=mean, fill=Species)) +
  geom_bar(stat="identity")
```
> **Exercise 8c.** Now, let's make a plot for ps_data. We want to visualize the mean `correct` for each `item` across each level of `condition`. Make sure the color of the bars represents the different `item`s. Instead of putting both levels of `condition` on the same plot, create two separate plots, one for each `condition`. Hint: use what you've learned so far about grouping and summarising; see *facet_wrap* above for help separating plots by `condition`!

```{r}


```