---
title: "R Bootcamp Day 2: Exploring the tidyverse"
author: "Cory Costello"
date: "9/26/2019"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

# Goals 

By the end of the day, you will:

+ How to read in and examine data
+ How to get values out of the rows and columns in your data
+ What a pipe is and how to use pipes to chain together `tidyverse` verbs
+ Using tidyverse "verbs" to modifiy and summarise data

# 0. Quick Refresher & Warm up 

Before we get started with new content, let's do a quick review of what we covered last time. 

## Exercise 0.1a
> Create a vector, x, and make it the numbers 1 throu 20. Multiply the 3rd element by 5.

```{r}
x <- 1:20
x[3]*5
```

## Exercise 0.1b
> Import the pragmatic scales dataset (pragmatic_scales_data.csv in the uoregon_r_bootcamp directory) as `ps_data`.

```{r}
library(rio)
ps_data <- import("pragmatic_scales_data.csv")
```

## Exercise 0.1c
> Use pipes to display the `correct` column of the ps_data for kids older than 2.

```{r}
library(tidyverse)
ps_data %>% 
  filter(age > 2) %>% 
  select(correct)
```

# 1. Diving deeper into dplyr

## 1.1 Refersher on select()
### 1.1.1 basic selectsd
```{r}
starwars %>% 
  select(name, homeworld)
```

### 1.1.2 selecting a range
```{r}
starwars %>% 
  select(name:homeworld)
```

### 1.1.3 selecting with helper functions

```{r}
starwars %>% 
  select(contains("color"))
```

### 1.1.4 Re-arranging columns with select

```{r}
starwars %>% 
  select(homeworld, name, everything())
```

## 1.2 Refersher on filter()

## 1.3 Transforming data with mutate() and transmute()

## 1.4 Grouping Data with group_by()

## 1.5 Summarizing data with summarize()

## 1.6 Brief Overview of Other dpyr verbs

### 1.6.1 arrange()

### 1.6.2 rename()

### 1.6.3 *_join()

### 1.6.4 case_when()

### 1.6.5 distinct()

### 1.6.6 count()

# 2. Cleaning & Reshaping Data with tidyr

A table is tidy if:
1. each  variable is its own colum, **and**
2. each row is a single observation or case

## 2.1 Lengthen with gather()

## 2.2 Widen with spread()

## 2.3 unite()

## 2.4 separate()

# 3. Factors with forcats

## 3.1 Inspecting Factors
### 3.1.1 fct_count()
### 3.1.2 fct_unique() 

## 3.2 Chaning the Order of Levels
### 3.2.1 fct_relevel()
### 3.2.2 fct_rev
### 3.2.3 fct_reorder 

## 3.3 Combining & Changing Factors
### 3.3.1 fct_c()
### 3.3.2 fct_recode()
### 3.3.2 fct_collapse()
### 3.3.3 fct_lump()

# 4. Manipulating strings with stringr
## 4.1 str_detect()

## 4.2 Changing case
### 4.2.1 str_to_lower()
### 4.2.2 str_to_upper()
### 4.2.3 str_to_title()

## 4.3 Changing content
### 4.3.1 str_replace()
### 4.3.2 str_pad()
### 4.3.3 str_trunc()

# 5. Introduction to ggplot2
## 5.1 ggplot basics
### 5.1.1 Grammar of Graphics
### 5.1.2 Aesthetic mapping


