---
title: "R Workshop Part 1"
author: "Cory K. Costello"
date: "3/30/2021"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---
# Goals 

By the end of this tutorial, you will know:

+ Basic `R` usage (using `R` as a calculator, creating variables, indexing)
+ Begin understanding tidyverse "grammar"
+ Begin understanding the grammar of data manipulation, `dplyr` 
+ What a pipe is and how to use pipes to chain together `tidyverse` verbs

The best way to do this tutorial is to walk through it slowly, executing each line and trying to understand what it does. You can execute a single line by placing your cursor somewhere in that line and hitting CTRL+enter on a PC or CMD+enter on a mac; you can execute a whole chunk at a time by hitting CTRL+shift+enter on a PC. 

# 1. Basic R Use

## 1.1. Basic Computing
R can be used as a calculator.

You can do basic arithmetic:

```{r}
# Add
2 + 3
# subtract
2 - 3
# multiplication
2 * 3
# Division
10/2
# Exponents
4^2
# Roots
4^(1/2)
```

Thankfully, R follows the order of operations (PEMDAS).

```{r}
(2^3)+4*(5/3)
```

These values aren't stored anywhere though.

## 1.2 Storing values in objects
To keep a value in memory, we need to assign it to an **object**. 

R has two assignment operators: `=` & `<-`. Let's use these to create an object called x, that is assigned the number 8. 

```{r}
x <- 8 
x = 8
```

Now we can call `x` by name; if we do this, it will print the value.
```{r}
x
```

Although `<-` and `=` both work, we'll strictly use `<-` from here on out. This brings up a general coding principal of the day:

*Style is important!* You want to use a consistent style so that others (including your future self) can easily and quickly read your code. I'll be using [Hadley Wickham's Style Guide](http://adv-r.had.co.nz/Style.html) from here on out, part of which is assigning variables with `<-`. Remember, the keyboard shortcut `Alt+-` or `Option+-` can be used to insert a `<-`.

Now, let's take a look at the variable `y`:

```{r eval = FALSE}
# Call Y
y 
```

D'oh! We didn't assign anything to `y`. Calling a variable that doesn't exist leads to an `Error`, 
```{r}
# Need to assign something to y!
y <- 2

y
```

Now, we can use these variables in calculations
```{r}
# add em up:
x + y
# multiply them
x * y
```

But remember, if we want to save *those* calculations, we need to assign them:
```{r}
# define the product as z
z <- x * y
```

## 1.3 Naming Variables
You may be wondering how to name variables in R. Well, there are just a few rules:

1. object names must start with a letter.
2. can contain alphanumeric characters, "_", and ".".
3. R is case sensitive, so `R` and `r` would be different variables.

This is another element of *style*. Using a consistent style when naming variables makes your life easier. 

Some options include:

+ snake_case
+ SCREAMING_SNAKE (*but why*)
+ camelCase
+ you.can.use.periods (*but don't!*)
  + `.` should be avoided, because it is used to define methods in R and you could end up conflicting with a method name.

I'll be using snake_case (again see [Hadley Wickham's Style Guide](http://adv-r.had.co.nz/Style.html)), but you can change to another option if you'd like.
```{r}
this_is_snake_case <- 8 # a_fine_choice
thisIsCamelCase <- 8 #alsoTotallyWorks
some.maniacs.use.periods <- 8 # probably.best.avoided
And_someTimes.People_doSTRANGEtHiNgS <- 8 # PLEASE NEVER DO ANYTHING LIKE THIS!
```

### Exercise 1.3a.

> Create a variable called var_1 that is equal to 4. Then create a variable called var_2 that is var_1 raised to the power of 6. Print the results.

```{r}

```

# 2. Types of Data

## 2.1. Types of Values
R has different types of data, and an object's type affects how it interacts with functions and other objects. So far, we've just been working with integers. R has several other types of data:

Type | Definition | Example
-----|------------|--------
Integer | whole numbers from -Inf to +Inf | 1, 2, -1, -10928
Numeric / Double | fractions & decimals from -Inf to +Inf | 1.3, .293764, -.1239217, `Inf`, `-Inf`
Character / String | quoted strings of letters, numbers, and allowed symbols| "1", "one", "o_n_e", "o-n-e", "o.n.e"
Logical / Boolean | logical constants of True or False | TRUE, FALSE, T, F
factor | ordered, labelled variable | variable for year in college labelled "Freshman", "Sophomore", etc.
There are a few special types worth knowing about too:

Type | Definition |Example
-----|------------|-------
NA   | Missing value; technically represented as different types but displayed as NA | NA, NA_character_, NA_integer_
NaN  | 'not a number' | 0/0
+Inf | positive infinity | Inf
-Inf | negative infinity | -Inf
NULL | The NULL object; an object that exists but is completely empty | NULL;

You can use `typeof()` to find out the type of a value or object:

```{r}
typeof(1)
typeof(TRUE)
typeof(1L)
typeof(NA_character_)
typeof(NA)
typeof(NULL)
typeof(NaN)
typeof(Inf)
typeof("one")
```

## 2.2 Vectors

Often, we're not working with individual values, but with multiple related values or a *vector* of values. 

### 2.2.1 Creating vectors

We can create a vector of ordered numbers using `starting_number:ending_number`. For example, we could make `x` a vector with the numbers between 1 and 5:

```{r}
x <- 1:5 # sequence of the form starting_number:ending_number
```

You'll notice that x has changed in the environment pane. Since it's a vector, it tells us what type of vector it is and it's length in addition to its contents (which can be abbreviated if the object is larger). In this case, it's a vector of integers (hence `int`) and is of length 5 (hence `[1:5]`).

We can also create a sequence using the function `seq()`. `seq()` is a function that creates a sequence of numbers. Like all R functions, you can learn more about it by accessing the help documentation (`?+function_name`). Let's take a look:

```{r eval = FALSE}
?seq
```

Interpreting R help documention is a skill itself, but a very worthwhile one.

What happens if we run `seq()` with no arguments?
```{r}
seq()
```

### 2.2.1.Q1 

> Why did this happen?**    

**A: **    

To make a sequence from 1 to 5 with this function, we have to set the `from =` and `to =` to those boundaries:

```{r}
seq(from = 1, to = 5)
```

We can also set one or more of the other arguments. The `by =` argument allows us to change the increment of the sequence. Let's say we wanted every other number between 1 and 5. We can do this by setting `by = 2`

```{r}
seq(from = 1, to = 5, by = 2)
```

We can also create a vector of any numbers we want using the `c()` function. You do this by putting numbers seperated by a `,` within the parentheses. 

```{r}
# We can also assign a vector of values this way
x <- c(2L, 8.5, 1, 9)
```

Note that it became a numeric vector (hence `num`), which is different from a vector of integers in that it allows decimal points. This brings us to an important feature of vectors:

Vectors are just 1-dimensional sequences **of a single type of data**. Because `1:5` was just integers, R stored the vector as `int`. Since we have `8.5`, all of the values will be numbers, even the data that could be integers. 

The general rule R uses is to set the vector to be the most permissive type necessary. 

Note that vectors can also include strings or character values.

```{r}
letters <- c("a", "b", "c", "d")
```

What happens if we put the vectors `x` and `letters` together? Let's take a look:

```{r}
mixed_vec <- c(x, letters)
mixed_vec
```

Notice the quotes? R turned all of our numbers into strings, since strings are more permissive than numbers. You can see that by running `typeof()`:

```{r}
typeof(mixed_vec)
```

This is called *coercion*. R coerces a vector into whichever type will accomodate all of the values. We can also coerce a vector ourselves using the `as.` family of functions. Let's coerce mixed_vec into a numeric:

```{r}
as.numeric(mixed_vec)
```


### Exercise 2.2.1a 

> Create an object called x that is assigned the number 8. Create an object called y that is a sequence of numbers from 2 to 16, by 2. Add x and y. What happens? 

```{r}

```

### Exercise 2.2.1b 

> Create an object called z that is a vector containing the numbers 5 and 10. Create an object called q that is a sequence of numbers from 5 to 25, by 5. Multiply q and z. What happens? 

```{r}

```

### Exercise 2.2.1c
> Create an object called a that is just the letter "a". Add a to x (from above). What happens? 

```{r eval = FALSE}

```

### Exercise 2.2.1d 
> Create a vector called b that is just the number 8 in quotes. Add b to x (from above). What happens? 

```{r}

```

```{r eval = FALSE}

```

### Exercise 2.2.1e 
> Find the type of each of the objects we just made (x, y, q, z, b)? 

```{r}

```

### Exercise 2.2.1f 
> Find some way to add b to x (hint: you may need to be coercive)? 

```{r}

```

### 2.2.2 Indexing vectors

The last thing to mention about vectors is that they are all indexed numerically in R, starting with 1. Since vectors are only 1 dimension, we index them with a single number in `[]`. Let's get just the first element of a new vector `z`:

```{r}
z <- 6:10
z[1]
```

What if we wanted the first two elements? We could do this by passing the sequence `1:2` as our index:

```{r}
z[1:2]
```

Or, we could get the first and third element by passing those indexes as a vector using `c()`:

```{r}
z[c(1, 3)]
```

We could also say which elements not to give us. Let's get rid of the first element:

```{r}
z[-1]
```

Or the first and third:
```{r}
z[-c(1, 3)]
```

Finally, if the elements in the vector have names, we can refer to them by name instead of their numerical index. You can see the names of a vector using `names()`

```{r}
names(z)
```

Looks like the elements in z have no names. We can change that by assigning them names using a vector of names:

```{r}
names(z) <- c("first", "second", "third", "fourth", "fifth")
```

And we can see the names:

```{r}
names(z) 
z
```

And use them to subset certain elements

```{r}
z["first"]
z[1]
```

#### 2.2.2.3 Manipulating elements in vectors

One thing that you can do with indexing is change elements within the vector. For example, we could change the first elemenet of `z` to missing, or `NA`:

```{r}
z[1] <- NA
z
```

You can also use indexing to transform a single element in a vector. Let's raise the second element to the power of 2:

```{r}
z[2] <- z[2]^2
z
```

Or use it to add to a vector. Let's add the number 11 as the sixth element of the vector z

```{r}
z[6] <- 11
z
```

And you can add a name for that element by referencing the index for the `names` attribute of `z`:

```{r}
names(z)[6] <- "sixth"
z
```

## Exercise 2.2.2a. 

> Create a vector called `named` that includes the numbers 1 to 5. Name each value a, b, c, d, and e (in order). Print the first element using numerical indexing and the last element using name indexes. 

```{r}

```

## Exercise 2.2.2b.
> Change the  third element of `named` to the value 21 and then show your results.

```{r}

```

## 2.3. Data Frames

The `data.frame` object is like a vector, but it's 2-dimensional and can contain heterogeneous data types. 

### 2.3.1 Creating Data Frames 

Note that we can create a dataframe from scratch with the `data.frame()` function:

```{r}
df_1 <- data.frame(c1 = c(1, 3),
                   c2 = c(2, 4), 
                   c3 = c("a", "b"),
                  row.names = c("r1", "r2"))
df_1
```

### 2.3.2 Indexing Data Frames

Like vectors, dataframes can be indexed by using numerical or named indices. However, dataframes require two indices in the [row, column] format. For example, we can use numerical indexing:

```{r}
df_1[1, 3]
```

or names: 

```{r}
df_1["r1", "c3"]
```

You can also get an entire row or column by leaving an index blank. Let's get all rows for column 2:

```{r}
df_1[, "c2"]
```

We can use the `$` in the form `data$column` (similar to `list$object` used for lists). Let's get the first column:

```{r}
df_1$c1
```

And, we can index a column using vector indexing, since a single column is (by definition) a vector (1-dimensional, with a single type of data). Let's get the first value in column 1:

```{r}
df_1$c1[1]
```

### 2.3.2.1 Modifying a Dataframe with Indexes

Just like lists, vectors, and matrices, you can modify a dataframe and add new elements or change exising elements by referencing indexes.

We could create c4 as the sum of c1 and c2:

```{r}
df_1$c4 <- df_1$c1 + df_1$c2
df_1
```

Or we could replace an element using indexing too. Let's replace c1 with c1^2:

```{r}
df_1$c1 =df_1$c1^2
df_1
```


### 2.3.3 Structure, Types, & Coercion in Dfs
How can we tell what type of data each column stores? We can use the `str()` function to get the *structure* of the data:

```{r}
str(df_1)
```

You can also coerce individual columns into other types. For example, we could turn c1 into an integer (from a numeric/double):

```{r}
df_1$c1 <- as.integer(df_1$c1)
df_1
```


### Exercise 2.3.2a 
> Make a data frame, called df_2, that has 3 columns.

c1 = 1, 2, 3
c2 = 2, 4, 6,
c3 = "a", "b", "c"

> After you create it, check the structure.

```{r}

```

### Exercise 2.3.2b 
>  Using df_2, add a fourth column, c4, which is the first and second columns multiplied together. 

```{r}

```

### Exercise 2.3.2c 
> Using df_2, try to add the first and second rows together. what happens? Why?

```{r}

```

# 3. Functions & Iteration

## 3.1. Basic Functions

Data (and objects more generally) are one of the building blocks of R. The other is functions. We've already used a handful of functions, including `seq()`, arithmetic functions (`+`, `*`, etc.), `matrix()`, `array()`, etc.

Functions take some form of an input, perform some operation, and then return some object(s) as output. Functions are made up of *arguments*. For example, let's take a look at the help documentation for `seq()` by copying ?seq into the console.

You can see it has the arguments `from`, `to`, `by`, `length.out`, and `along.with`. You might also notice that each of the arguments have a value after the `=` in the documentation. These values are the *defaults*; they are what the arguments will be set to if you don't specify them. In fact, since all of the arguments have defaults, we don't have to specify any to run `seq()` as we saw earlier today.

```{r}
seq()
```

Let's take a look at a new function, `mean()`. This is used to get the mean of a set of numbers. What happens if we run it without any arguments?

```{r eval = FALSE}
?mean()
```

We get an error telling us that the argument "x" is missing and has no default. Whenever you see this error, it means you are missing a required argument (i.e., an argument without a default). If we look at the documention, you can see x is the data from which to calculate a mean.

Let's see what happens when we run `mean()` on the matrix `mat` we created eariler:

```{r}
mean(df_1$c1)
```

Note that `mean()` has two more optional arguments listed: `trim`, which returns a trimmed mean, and `na.rm` which takes a logical value indicating if it should remove missing values or not before it calculates the mean (set to `FALSE` by default). What happens if we don't remove NAs before calculating the mean? Let's check it out.

```{r}
vec_w_na <- c(1, 2, 3, 4, 5, 6, NA, 2, 4)
mean(vec_w_na)
```

It returns `NA`. This is important to remember: NAs are contagious! One in a vector will cause many functions to return NA (unless they remove them by default). This sort of makes sense - the mean of `vec_w_na` in its entirity is unknown, since we don't know what the `NA` value is. That's why you have to remove them by setting `na.rm = TRUE`

```{r}
mean(vec_w_na, na.rm = TRUE)
```

`sd()` can be used to get the standard deviation and is structured very similarly:

```{r}
sd(vec_w_na, na.rm = TRUE)
```

You can get the length of many objects with `length()`:

```{r}
length(vec_w_na)
```

`nrow()` and `ncol()` can be used to get the number of rows or columns in a matrix or data frame:

```{r}
nrow(df_1)
ncol(df_1)
```

## 3.2. Arguments are Positional

Take a look at the `sd()` function documentation (`?sd`). Notice that there are two arguments and they are in order, `x` followed by `na.rm = FALSE`. You can set arguments by name:

```{r}
sd(x = vec_w_na, na.rm = TRUE)
```

You can also set them positionally:

```{r}
sd(vec_w_na, TRUE)
```

When using arguments positionally (without their names), you **need to make sure the arguments are in the right order.** If you name the arguments, you can actually put them in a different order:

```{r}
sd(na.rm = TRUE, x = vec_w_na)
```

Now let's try the `mean()` function:

```{r eval = FALSE}
mean(vec_w_na, TRUE)
```

### 3.2.Q1: 
> What happenned?     
**A: **

```{r}
mean(vec_w_na, na.rm = TRUE) # or...
mean(vec_w_na, 0, TRUE)
```

## 3.3. Packages

So far, we've been working with functions that are already installed and loaded when we open R. However, many of the functions we want to use are not part of the basic R install. They come in `packages` that other R users make and share.

### 3.3.1 Installing Packages

The most common way to get a package is to download it from [CRAN](https://cran.r-project.org/) using the code `install.packages()` function. Packages on CRAN have to pass certain tests, so any package on CRAN should be safe and basically work (there can be bugs).

One package we're going to use is the package `rio`, which has really easy functions for importing and exporting data. To install a package, you want to run `install.packages("package_name")`.

```{r eval = FALSE}
install.packages("rio")
```

### 3.3.2 Updating Packages

You occasionally need to update packages, which you can do with `update.packages()`:

```{r eval = FALSE}
update.packages("rio")
```

### 3.3.3 Opening Packages

Installing a package puts a copy of it into our personal library that R has access to. If we want to use it, we have to open the package in our working session. We do this with the `library()` function:

```{r}
library(rio)
library("rio") # also works
```

Now that we've opened `rio`, we have access to its functions like `import()` and `export()`.

What happenes when we open a package? Loading a package basically makes the contents of it searchable by R. This makes it so it can find the functions a particular package has. You can see that by running the `search()` command:

```{r}
search()
```

Let's install another package we're going to use quite a bit, which is the `tidyverse`. The tidyverse isn't exactly a package, it's a set of 8 packages and a broader coding framework for R. We'll get into that a bit at the end of today and for the next two sessions, but let's get the package downloaded now:

```{r eval = FALSE}
install.packages("tidyverse")
```

And let's load it:

```{r}
library(tidyverse)
```

### 3.3.4 Conflicts

You'll notice that when we load the tidyverse, it tells us the 8 different packages it loads. It also tells us that we have conflicts with other packages.

Conflicts occur when the same name is used for different things. For example, the `dplyr` package and the `stats` package (preloaded package) both have a function called `filter()`. When we call `filter()`, R will only call one of those functions and it might not be the one we want. 

Which one will R choose? R has an order in which it searches. It starts with the Global Environment, then searches packages in the order that they were loaded, searching more recently loaded packages first. You can actually see the order by using the `search()` command:

```{r}
search()
```

As you can see, the tidyverse packages come before rio, which comes before the pre-loaded packages (rstudio, stats, etc.).

The last thing I want to mention about packages is that you can tell R explicitly that you want an object from a particular package using the notation `package::object`, or more commonly `package::function()`. For example, run the two pieces of code below:

```{r eval = FALSE}
?stats::filter
?dplyr::filter
```

This can be useful if you have conflicting packages and want to make sure you're calling the right function.


